---
title: "Exploratory Analysis"
output: github_document
---

```{r setup}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Load the weather data

```{r}
weather_df =
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"),
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>% 
  mutate(
    name = recode(
      id,
      USW00094728 = "CentralPark_NY",
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>% 
  select(name, id, everything())

weather_df
```

lubridate::floor_date   code rounds month and date to the lowest.

## `group_by`

```{r}
weather_df %>% 
  group_by(name)
```

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  ungroup()
```

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  ungroup(month)
```

## counting things

count month observations
```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n())
```

```{r}
weather_df %>% 
  group_by(name) %>% 
  summarize(n_obs = n())
```

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(n_obs = n())
```

we can use `count()` as well - will generate the same thing as group_by()

```{r}
weather_df %>% 
  count(month, name = "n_obs")
```

```{r}
weather_df %>% 
  count(name, month, name = "n_obs")
```

**Never** use base R's `table`

```{r, eval = FALSE}
weather_df %>% 
  pull(month) %>% 
  table()
```

Why?
  Because the result of this is not a dataframe.  Not useable for later purpose
  
Other helpful counters

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    n_obs = n(),
    n_days = n_distinct(date)
    )
```
This above code tells me the number of observations in the grouping specificed (month), AND it also tells us how many of those dates are distinct.
