---
title: "Visualization"
subtitle: "Part II"
output: github_document
---

```{r setup}
library(tidyverse)
library(patchwork)
```

## Load the weather data

```{r}
weather_df =
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"),
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>% 
  mutate(
    name = recode(
      id,
      USW00094728 = "CentralPark_NY",
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>% 
  select(name, id, everything())

weather_df
```

## remember this plot ..?

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5)
```

## Labels

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maximum daily temperature (C)",
    caption = "Data from rnoaa package; temperatures in 2017."
  )
```


## Scales - helps control x-axis and y-axis properties

Start with the same plot; can manipulate x and y scales.

Depends on type of axis: continuous variable vs. discrete variable, for example.  make sure to add the + (not pipe) between labs() and scale_x_continuous().
trans means transformation, 
 e.g. trans = "sqrt"
 e.g. trans = "log"
position means change position of y axis,
 e.g. position = "right"

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maximum daily temperature (C)",
    caption = "Data from rnoaa package; temperatures in 2017."
  ) +
  scale_x_continuous(
    breaks = c(-15, 0, 15),
    labels = c("-15 C", "0", "15")
  ) +
  scale_y_continuous(
    position = "right"
  )
```

Look at color scales
 e.g. scale_color_hue(h = c(0, 360))   0 to 360 is range.
 e.g. name = "Location"   will change the legend for the name of locations
viridis::scale_color_viridis() is a diffrent color package.  Assumes continuous variable so must define, in this case, a discrete variable by specifying "discrete = TRUE"

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maximum daily temperature (C)",
    caption = "Data from rnoaa package; temperatures in 2017."
  ) +
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  )
```

## Themes

Shift the legend.

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maximum daily temperature (C)",
    caption = "Data from rnoaa package; temperatures in 2017."
  ) +
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE) +
  theme(
    legend.position = "bottom"
  )
```

Change the overall theme - should go last in order to not overwrite other specifications
theme_bw()
theme_minimal()
theme_classic()
ggthemes::theme_economist()
ggthemes::theme_excel()

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maximum daily temperature (C)",
    caption = "Data from rnoaa package; temperatures in 2017."
  ) +
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## Setting options

```{r, eval = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Data args in `geom`

```{r}
central_park =
  weather_df %>% 
  filter(name == "CentralPark_NY")

waikiki =
  weather_df %>% 
  filter(name == "Waikiki_HA")

ggplot(data = waikiki, aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_line()

```

if I want to superimpose central park data, you can use the "data =" argument.

```{r}
central_park =
  weather_df %>% 
  filter(name == "CentralPark_NY")

waikiki =
  weather_df %>% 
  filter(name == "Waikiki_HA")

ggplot(data = waikiki, aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_line(data = central_park)

```

## `patchwork`

remember faceting?

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, fill = name)) +
  geom_density(alpha = .5) +
  facet_grid(. ~ name)
```

what happens when you want multipanel plots but can't facet ...?
that's where patchwork comes in

```{r}
tmax_tmin_plot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  theme(legend.position = "none")

precip_dens_p =
  weather_df %>% 
  filter(prcp > 0) %>% 
  ggplot(aes(x = prcp, fill = name)) +
  geom_density(alpha = .5) +
  theme(legend.position = "none")

tmax_date_p =
  weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none")
```

So now you have these three plots that have fundamental differences.  you can't use facet to organize and present together
But, you can use patchwork.  Can do a lot with this - go to patchwork webpage to learn more

```{r}
tmax_tmin_plot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  theme(legend.position = "none")

prcp_dens_p =
  weather_df %>% 
  filter(prcp > 0) %>% 
  ggplot(aes(x = prcp, fill = name)) +
  geom_density(alpha = .5) +
  theme(legend.position = "none")

tmax_date_p =
  weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none")

tmax_tmin_plot + prcp_dens_p
tmax_tmin_plot + prcp_dens_p + tmax_date_p
tmax_tmin_plot + (prcp_dens_p + tmax_date_p)
tmax_tmin_plot / (prcp_dens_p + tmax_date_p)
(tmax_tmin_plot + prcp_dens_p) / tmax_date_p
```

## the challenge of data manipulation coupled with ggplot - this is a next level interaction - THIS IS HARD.  THIS IS A LOT.  DON'T GET DISCOURAGED THAT THIS ISN'T EASY!

## Data manipulation

Control your factors.

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha = .5)
```

R, behind the scenes, has defaulted chr variation in alphabetical order (as a factor variable)
We are trying to now change into a different order by manipulating the code (not a ggplot customization issue)
Data manipulation issue/challenge now dealing with factor variable-type

```{r}
weather_df %>% 
  mutate(
    name = factor(name)
  ) %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha = .5)

weather_df

```

This has now changed name to a factor.  The plot looks the same

```{r}
weather_df %>% 
  mutate(
    name = factor(name),
    name = forcats::fct_relevel(name, c("Waikiki_HA"))
  ) %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha = .5)
```

Now, behind the scenes in that previous chunk, weather_df name variable is still character but it has changed the factor leveling to make Waikiki first, CentralPark goes to second, Waterhole still third.

There are other relevel commands and customizations.

Now,
What if I wanted densities for tmin and tmax simultaneously?  Make tmax and tmin long format.

```{r}
weather_df %>% 
  filter(name == "CentralPark_NY") %>% 
  pivot_longer(
    tmax:tmin,
    names_to = "observation",
    values_to = "temperatures"
  ) %>% 
  ggplot(aes(x = temperatures, fill = observation)) +
  geom_density(alpha = .5)
```

Now facet by name

```{r}
weather_df %>% 
  pivot_longer(
    tmax:tmin,
    names_to = "observation",
    values_to = "temperatures"
  ) %>% 
  ggplot(aes(x = temperatures, fill = observation)) +
  geom_density(alpha = .5) +
  facet_grid(. ~ name)
```

The point here is sometimes it is not a ggplot issue - it is a data tidiness or data manipulation issue.  In the preceding example, the pivot longer concept was essential.

## Revisit the pups

Drive home the interconnectedness between the data import/manipulation and plots.

Data from the FAS study.

```{r}
pups_data =
  read_csv("./data/FAS_pups.csv") %>% 
  janitor::clean_names() %>% 
  mutate(sex = recode(sex, `1` = "male", `2` = "female"))

litters_data =
  read_csv("./data/FAS_litters.csv") %>% 
  janitor::clean_names() %>% 
  separate(group, into = c("dose", "day_of_tx"), sep = 3)

fas_data = left_join(pups_data, litters_data, by = "litter_number")

fas_data %>% 
  ggplot(aes(x = dose, y = pd_ears)) +
  geom_violin() +
  facet_grid(. ~ day_of_tx)

fas_data %>% 
  ggplot(aes(x = dose, y = pd_ears)) +
  geom_violin() +
  facet_grid(. ~ day_of_tx)

```

What is you wanted to look at ears, eyes, pivot, and walk all at the same time?  The all deal with milestones related to postnatal day.

```{r}
fas_data %>% 
  select(dose, day_of_tx, starts_with("pd_")) %>% 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome",
    values_to = "pn_day"
  ) %>% 
  drop_na() %>% 
  mutate(outcome = forcats::fct_relevel(outcome, "pd_ears", "pd_pivot", "pd_walk", "pd_eyes")) %>% 
  ggplot(aes(x = dose, y = pn_day)) +
  geom_violin() +
  facet_grid(day_of_tx ~ outcome)
```

This previous example is a complex process of data manipulation and plotting

Learning Assessment #1

Revisit the plot showing tmax against date for each location.  Use labels, scale options, and theme changes to improve the readibility of this plot.

```{r}
weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_smooth(se = FALSE) +
  geom_point(aes(size = prcp), alpha = .75) +
  labs(
    title = "Temperature plot",
    x = "Date",
    y = "Maximum daly temperature (C)",
    caption = "Data from the rnoaa package"
  ) +
  viridis::scale_color_viridis(discrete = TRUE) +
  theme_minimal() +
  theme(legend.position = "bottom")
  
```

